Miniproject - Elliptic Curves

### 2022-03-16 Elliptic Curves/submission.tex
Did 4i

\newtheorem{prop}{Proposition}
\newtheorem{theorem}[prop]{Theorem}
\newtheorem{lemma}[prop]{Lemma}
\newtheorem{corollary}[prop]{Corollary}

\theoremstyle{definition}
\newtheorem{problem}[prop]{Problem}
\newtheorem{alg}[prop]{Algorithm}
\newtheorem{definition}[prop]{Definition}
\newtheorem{example}[prop]{Example}
\newtheorem{remark}[prop]{Remark}

@@

    Note that the usual approach to bound the size of $E_{\mathrm{tors}}(\Q)$ is use to the theorem that this embeds into $\tilde{E}(\F_p)$ whenever $\tilde{E}$ is an Elliptic Curve.
    However, for this example, this was not sufficient, as we could not find a prime such that the group $\Z/4\Z \oplus \Z/2\Z$ does not embed into $\tilde{E}(\F_p)$.
    In the next part, we want to study this phenomenon more carefully and indeed see that there is no such prime, i.e. it is impossible to show that $\#E_{\mathrm{tors}}(\Q) \neq 8$ by just considering the reductions modulo $p$.
\end{example}
First, it is convenient to have a closed formula for the $x$-coordinate of $[2]P$ for a point $P$ on an Elliptic Curve.
\begin{prop}
    \label{prop:duplication_formula}
    Let $E: y^2 = x^3 + a_2 x^2 + a_4 x + a_6$ be an Elliptic Curve over a field $k$.
    For a point $P \in E$ with $P \neq \O$ denote by $x(P)$ its (affine) $x$-coordinate.
    Then have for all $P \in E$ with $P \neq -P$ that
    \begin{equation*}
        x([2]P) = \frac {x(P)^4 - 2a_4 x(P)^2 - 8a_6x(P) + a_4^2 - 4a_2a_6} {4(x(P)^3 + a_2 x(P)^2 + a_4 x(P) + a_6)}
    \end{equation*}  
\end{prop}
\begin{prop}
    Let $E: y^2 = x(x + 1)(x - 8)$ be the Elliptic Curve from the previous example.
    The for each prime $p \geq 5$, have that $\Z/4\Z \oplus \Z/2\Z$ is a subgroup of $\tilde{E}(\F_p)$.
\end{prop}
\begin{proof}
    First of all, note that the duplication formula from Proposition~\ref{prop:duplication_formula} has the form
    \begin{equation*}
        x([2]P) = \frac {x(P)^4 + 16 x(P) + 64} {4 x(P)^3 - 28 x(P)^2 - 32 x(P)}
    \end{equation*}
    Consider any prime $p \geq 5$.

@@

\section*{Question 3}
\begin{lemma}
    \label{prop:relation_y_uv}
    Let $E: y^2 = x^3 + a_2 x^2 + a_4 x + a_6$ be an Elliptic Curve defined over $k$.
    Suppose that $E$ has a nontrivial 5-torsion point $P = (\alpha_1, \beta_1) \in E(k)$.
    wlog $\alpha_1 = 0$.
    Let further $(\alpha_2, \beta_2) = [2](0, \beta)$.
    Then
    \begin{equation*}
        \alpha_2^3 = -4\beta\beta_2
    \end{equation*}
\end{lemma}
\begin{proof}
    Consider the following SAGE script:
    \lstinputlisting[language = python]{./3i_algebra.sage}
    The output shows that $\alpha_2^3 = -4 \beta \beta_2$ already follows from the facts
    \begin{align*}
        &(0, \beta) \in E, \\
        &(\alpha_2, \beta_2) \in E, \\
        &[2](0, \beta) = (\alpha_2, \beta_2), \\
        &[2](\alpha_2, \beta_2) = (0, -\beta)
    \end{align*}
    which hold by assumption that $(0, \beta)$ has order $5$.
    The claim follows.
\end{proof}
\begin{lemma}
    Let
    \begin{equation*}
        E: y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2 + a_4 x + a_6    
    \end{equation*}
    and
    \begin{equation*}
        E': y^2 + a'_1 x y + a'_3 y = x^3 + a'_2 x^2 + a'_4 x + a'_6    
    \end{equation*}
    be Elliptic Curves.
    If there are five distinct points $P_1, ..., P_5 \in E \cap E'$ in ``sufficiently independent position'', then $E = E'$ (meaning $a_i = a'_i$).

@@

    Under $\psi$, we now see that
    \begin{equation*}
        (0, \ \beta) \mapsto (0, u^3 \beta + t) = \left( 0, \ \frac {\beta} {2\beta_2} - \frac {\beta} {2\beta_2} \right) = (0, 0) \in E'
    \end{equation*}
    and
    \begin{align*}
        (\alpha_2, \ \beta_2) \mapsto (u^2 \alpha_2, \ u^3 \beta_2 + u^2 s \alpha_2 + t) &= \left(\frac {\alpha_2^3} {4\beta^2}, \ \frac {\beta_2^2} {2\beta^2} + \frac {\beta_2} {2\beta} \left( \frac {\beta_2} {\beta} + 1 \right) - \frac {\beta_2} {2\beta} \right) \\
        &= \left( -\frac {\beta_2} {\beta}, \ \frac {\beta_2^2} {\beta^2} \right) = (-v, \ v^2) \in E'
    \end{align*}
    and
    \begin{align*}
        (\alpha_2, \beta_2) \mapsto (u^2 \alpha_2, \ -u^3 \beta_2 + u^2 s \alpha_2 + t) &= \left(\frac {\alpha_2^3} {4\beta^2}, \ -\frac {\beta_2^2} {2\beta^2} + \frac {\beta_2} {2\beta} \left( \frac {\beta_2} {\beta} + 1 \right) - \frac {\beta_2} {2\beta} \right) \\
        &= ( -\frac {\beta_2} {\beta}, \ 0) = (-v, \ 0) \in E'
    \end{align*}
    and
    \begin{align*}
        (0, \ -\beta) \mapsto (0, -u^3 \beta + t) = \left( 0, \ -\frac {\beta_2} {2\beta} - \frac {\beta_2} {2\beta} \right) = (0, \ -v) \in E'
    \end{align*}
    and of course $\O \mapsto \O$.
    Hence we have $\#(E' \cap E'') \geq 5$ and by the previous proposition we see that $E' = E''$.
\end{proof}
\begin{corollary}
    Let $E$ be an Elliptic Curve defined over $\Q$.
    Then $E(\Q)$ has a nontrivial 5-torsion point if and only if $E$ is isomorphic (over $\Q$) to an Elliptic Curve given by an equation of the form
    \begin{equation*}
        y^2 + (v + 1)x y + v y = x^3 + v x^2
    \end{equation*}
\end{corollary}
In the lecture, it was mentioned that a theorem of Mazur states that the torsion group of an Elliptic Curve $E$ defined over $\Q$ has one of the following forms
\begin{itemize}
    \item $\Z/n\Z$ for $n \in \{ 1, ..., 10, 12 \}$
    \item $\Z/n\Z \oplus \Z/2\Z$ for $n \in \{ 2, 4, 6, 8 \}$
\end{itemize}
Hence, we are quite limited for $N$ a similar idea can work.
First, note that the case $N = 2$ is easy.
\begin{prop}
    Let $E$ be an Elliptic Curve defined over $\Q$.
    Then $E(\Q)$ has a nontrivial 2-torsion point if and only if $E$ is isomorphic (over $\Q$) to an Elliptic Curve given by an equation of the form
    \begin{equation*}
        y^2 = x^3 + a_2 x^2 + a_4 x
    \end{equation*}
\end{prop}
\begin{proof}
    Assume $E: y^2 = x^3 + a_2 x^2 + a_4 x$ is an Elliptic Curve.
    Then clearly $(0, 0) \in E(\Q)$ and $(0, 0) = -(0, 0)$, so $(0, 0)$ is nontrival 2-torsion point.
    Hence, there is a nontrivial 2-torsion point in $E'(\Q)$ for all Elliptic Curves $E'$ that are isomorphic (over $\Q$) to $E$.

@@

    Now consider the line through $\O$ and $(0, 0)$, which is given by $x = 0$.
    Plugging this in yields $y^2 + v y = 0$ and so the third point of intersection is $(0, -v)$.
    Now note that $(0, 0)$, $(0, -v)$ and $\O$ are colinear, so $(0, 0) + (0, -v) + \O = \O$, hence $(0, 0) = -[2](0, 0)$ has order 3.
\end{proof}
\end{document}

>>>>

\theoremstyle{definition}

\newtheorem{prop}[subsection]{Proposition}
\newtheorem{theorem}[subsection]{Theorem}
\newtheorem{lemma}[subsection]{Lemma}
\newtheorem{corollary}[subsection]{Corollary}

\newtheorem{problem}[subsection]{Problem}
\newtheorem{alg}[subsection]{Algorithm}
\newtheorem{definition}[subsection]{Definition}
\newtheorem{example}[subsection]{Example}
\newtheorem{remark}[subsection]{Remark}

@@

    Note that the usual approach to bound the size of $E_{\mathrm{tors}}(\Q)$ is use to the theorem that this embeds into $\tilde{E}(\F_p)$ whenever $\tilde{E}$ is an Elliptic Curve.
    However, for this example, this was not sufficient, as we could not find a prime such that the group $\Z/4\Z \oplus \Z/2\Z$ does not embed into $\tilde{E}(\F_p)$.
    In the next part, we want to study this phenomenon more carefully and indeed see that there is no such prime, i.e. it is impossible to show that $\#E_{\mathrm{tors}}(\Q) \neq 8$ by just considering the reductions modulo $p$.
\end{example}
First, it is convenient to have a closed formula for the $x$-coordinate of $[2]P$ for a point $P$ on an Elliptic Curve.
\begin{prop}[Duplication Formula]
    \label{prop:duplication_formula}
    Let $E: y^2 = x^3 + a_2 x^2 + a_4 x + a_6$ be an Elliptic Curve over a field $k$.
    For a point $P \in E$ with $P \neq \O$ denote by $x(P)$ its (affine) $x$-coordinate.
    Then have for all $P \in E$ with $P \neq -P$ that
    \begin{equation*}
        x([2]P) = \frac {x(P)^4 - 2a_4 x(P)^2 - 8a_6x(P) + a_4^2 - 4a_2a_6} {4(x(P)^3 + a_2 x(P)^2 + a_4 x(P) + a_6)}
    \end{equation*}  
\end{prop}
\begin{prop}
    Let $E: y^2 = x(x + 1)(x - 8)$ be the Elliptic Curve from the previous example.
    The for each prime $p \geq 5$, have that $\Z/4\Z \oplus \Z/2\Z$ is a subgroup of $\tilde{E}(\F_p)$.
\end{prop}
\begin{proof}
    First of all, note that the duplication formula from Proposition~\ref{prop:duplication_formula} has the form
    \begin{equation*}
        x([2]P) = \frac {x(P)^4 + 16 x(P) + 64} {4 x(P)^3 - 28 x(P)^2 - 32 x(P)}
    \end{equation*}
    Consider any prime $p \geq 5$.

@@

\section{Question 3}
\begin{lemma}
    \label{prop:relation_y_uv}
    Let $E: y^2 = x^3 + a_2 x^2 + a_4 x + a_6$ be an Elliptic Curve defined over $k$.
    Suppose that $E$ has a nontrivial 5-torsion point $P = (\alpha_1, \beta_1) \in E(k)$.
    wlog $\alpha_1 = 0$.
    Let further $(\alpha_2, \beta_2) = [2](0, \beta)$.
    Then
    \begin{equation*}
        \alpha_2^3 = -4\beta\beta_2
    \end{equation*}
\end{lemma}
\begin{proof}
    Consider the following SAGE script:
    \lstinputlisting[language = python]{./3i_algebra.sage}
    The output shows that $\alpha_2^3 = -4 \beta \beta_2$ already follows from the facts
    \begin{align*}
        &(0, \beta) \in E, \\
        &(\alpha_2, \beta_2) \in E, \\
        &[2](0, \beta) = (\alpha_2, \beta_2), \\
        &[2](\alpha_2, \beta_2) = (0, -\beta)
    \end{align*}
    which hold by assumption that $(0, \beta)$ has order $5$.
    The claim follows.
\end{proof}
\begin{lemma}
    Let
    \begin{equation*}
        E: y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2 + a_4 x + a_6    
    \end{equation*}
    and
    \begin{equation*}
        E': y^2 + a'_1 x y + a'_3 y = x^3 + a'_2 x^2 + a'_4 x + a'_6    
    \end{equation*}
    be Elliptic Curves.
    If there are five distinct points $P_1, ..., P_5 \in E \cap E'$ in ``sufficiently independent position'', then $E = E'$ (meaning $a_i = a'_i$).

@@

    Under $\psi$, we now see that
    \begin{equation*}
        (0, \ \beta) \mapsto (0, u^3 \beta + t) = \left( 0, \ \frac {\beta} {2\beta_2} - \frac {\beta} {2\beta_2} \right) = (0, 0) \in E'
    \end{equation*}
    and
    \begin{align*}
        (\alpha_2, \ \beta_2) \mapsto (u^2 \alpha_2, \ u^3 \beta_2 + u^2 s \alpha_2 + t) &= \left(\frac {\alpha_2^3} {4\beta^2}, \ \frac {\beta_2^2} {2\beta^2} + \frac {\beta_2} {2\beta} \left( \frac {\beta_2} {\beta} + 1 \right) - \frac {\beta_2} {2\beta} \right) \\
        &= \left( -\frac {\beta_2} {\beta}, \ \frac {\beta_2^2} {\beta^2} \right) = (-v, \ v^2) \in E'
    \end{align*}
    and
    \begin{align*}
        (\alpha_2, \beta_2) \mapsto (u^2 \alpha_2, \ -u^3 \beta_2 + u^2 s \alpha_2 + t) &= \left(\frac {\alpha_2^3} {4\beta^2}, \ -\frac {\beta_2^2} {2\beta^2} + \frac {\beta_2} {2\beta} \left( \frac {\beta_2} {\beta} + 1 \right) - \frac {\beta_2} {2\beta} \right) \\
        &= ( -\frac {\beta_2} {\beta}, \ 0) = (-v, \ 0) \in E'
    \end{align*}
    and
    \begin{align*}
        (0, \ -\beta) \mapsto (0, -u^3 \beta + t) = \left( 0, \ -\frac {\beta_2} {2\beta} - \frac {\beta_2} {2\beta} \right) = (0, \ -v) \in E'
    \end{align*}
    and of course $\O \mapsto \O$.
    Hence we have $\#(E' \cap E'') \geq 5$ and by the previous proposition we see that $E' = E''$.
\end{proof}
\begin{corollary}[3(i)]
    Let $E$ be an Elliptic Curve defined over $\Q$.
    Then $E(\Q)$ has a nontrivial 5-torsion point if and only if $E$ is isomorphic (over $\Q$) to an Elliptic Curve given by an equation of the form
    \begin{equation*}
        y^2 + (v + 1)x y + v y = x^3 + v x^2
    \end{equation*}
\end{corollary}
\begin{proof}
    Note that $E$ is isomorphic (over $\Q$) to an Elliptic Curve defined by a Weierstraß equation, so assume wlog
    \begin{equation*}
        E: y^2 = x^3 + a_2 x^2 + a_4 x + a_6
    \end{equation*}
    Now let $P = (\alpha, \beta)$ be a nontrivial 5-torsion point.
    Then $E$ is isomorphic via $(x, y) \mapsto (x - \alpha, \beta)$ to an Elliptic Curve
    \begin{equation*}
        E': y^2 = x^3 + a'_2 x^2 + a'_4 x
    \end{equation*}
    and $(0, \beta)$ is a nontrivial 5-torsion point of $E'$.
    The claim follows now by Proposition~\ref{prop:characterization_5_torsion}.

    Conversely, assume that
    \begin{equation*}
        E: y^2 + (v + 1) x y + v y = x^3 + v x^2
    \end{equation*}
    For example by the duplication formula~\ref{prop:duplication_formula} find that
    \begin{equation*}
        [2](0, 0) = (-v, v^2), \quad [2](-v, v^2) = (0, -v)
    \end{equation*}
    and thus $[4](0, 0) + (0, 0) = \O$, hence $(0, 0)$ has order 5.
\end{proof}
In the lecture, it was mentioned that a theorem of Mazur states that the torsion group of an Elliptic Curve $E$ defined over $\Q$ has one of the following forms
\begin{itemize}
    \item $\Z/n\Z$ for $n \in \{ 1, ..., 10, 12 \}$
    \item $\Z/n\Z \oplus \Z/2\Z$ for $n \in \{ 2, 4, 6, 8 \}$
\end{itemize}
Hence, we are quite limited for $N$ a similar idea can work.
First, note that the case $N = 2$ is easy.
\begin{prop}[3(ii)]
    Let $E$ be an Elliptic Curve defined over $\Q$.
    Then $E(\Q)$ has a nontrivial 2-torsion point if and only if $E$ is isomorphic (over $\Q$) to an Elliptic Curve given by an equation of the form
    \begin{equation*}
        y^2 = x^3 + a_2 x^2 + a_4 x
    \end{equation*}
\end{prop}
\begin{proof}
    Assume $E: y^2 = x^3 + a_2 x^2 + a_4 x$ is an Elliptic Curve.
    Then clearly $(0, 0) \in E(\Q)$ and $(0, 0) = -(0, 0)$, so $(0, 0)$ is nontrival 2-torsion point.
    Hence, there is a nontrivial 2-torsion point in $E'(\Q)$ for all Elliptic Curves $E'$ that are isomorphic (over $\Q$) to $E$.

@@

    Now consider the line through $\O$ and $(0, 0)$, which is given by $x = 0$.
    Plugging this in yields $y^2 + v y = 0$ and so the third point of intersection is $(0, -v)$.
    Now note that $(0, 0)$, $(0, -v)$ and $\O$ are colinear, so $(0, 0) + (0, -v) + \O = \O$, hence $(0, 0) = -[2](0, 0)$ has order 3.
\end{proof}

\section{Question 4}

\begin{example}
    The Elliptic Curve $E: y^2 = x(x + 6 + 1)$ has rank $0$.
\end{example}
\begin{proof}
    As in the lecture, consider
    \begin{align*}
        &E': y^2 = x(x^2 - 12x + 32) \\
        &\phi: E \to E', \quad (u, v) \mapsto \left( \frac {y^2} {x^2}, \ y \frac {x^2 - 1} {x^2} \right) \\
        &\hat{\phi}: E' \to E, \quad (u, v) \mapsto \left( \frac {y^2} {4x^2}, \ y \frac {x^2 - 1} {8x^2} \right) \\
        &q: E'(\Q)/\phi(E(\Q)) \to \Q^*/S, \quad \overline{(u, v)} \mapsto \begin{cases}
            \overline{u} & \text{if $u \neq 0$} \\
            \overline{32} & \text{if $u = 0$}
        \end{cases} \\
        &\hat{q}: E(\Q)/\phi(E'(\Q)) \to \Q^*/S, \quad \overline{(u, v)} \mapsto \begin{cases}
            \overline{u} & \text{if $u \neq 0$} \\
            \overline{1} & \text{if $u = 0$}
        \end{cases}
    \end{align*}
    \paragraph{Find $E'(\Q)/\phi(E(\Q))$} As in the lecture, consider $r \divides 32$ square-free, i.e. $r \in \{ \pm 1, \pm 2 \}$.

    For $r = 2$, have that $(l, m, n) = (2, 1, 0)$ solves
    \begin{equation*}
        2l^4 - 12l^2m^2 + 16m^4 = n^2
    \end{equation*}
    and indeed we find $(8, 0) \in E'(\Q)$.

    For $r = -1$, note that
    \begin{equation*}
        -l^4 - 12l^2m^2 - 32m^4 = n^2
    \end{equation*}
    has no nontrivial solutions in $\Q$, as the left-hand side is always $\leq 0$ and the right-hand side is $\geq 0$.

    Since $-2 = -1 \cdot 2$, we see that $\im(q) = \langle 2 \rangle$ and $E'(\Q)/\phi(E(\Q)) = \langle (8, 0) \rangle$.

    \paragraph{Find $E(\Q)/\hat{\phi}(E'(\Q))$} Consider $r \divides 1$, i.e. $r \in \{ \pm 1 \}$.

    For $r = -1$, have that $(l, m, n) = (1, 1, 2)$ solves
    \begin{equation*}
        -l^4 + 6l^2m^2 - m^2 = n^2
    \end{equation*}
    and indeed find $(-1, 2) \in E(\Q)$.

    Thus find $\im(\hat{q}) = \langle -1 \rangle$ and $E(\Q)/\hat{\phi}(E'(\Q)) = \langle (-1, 2) \rangle$.

    \paragraph{Find the rank of $E$} By the above two steps, we see that
    \begin{equation*}
        E(\Q)/[2]E(\Q) = \langle (-1, 2), \hat{\phi}((8, 0)) \rangle = \langle (-1, 2), (0, 0) \rangle
    \end{equation*}
    Now observe that $[2](-1, 2) = (0, 0)$ and $[2](0, 0) = \O$.
    Hence $E(\Q) = E_{\mathrm{tors}}(\Q)$ and the rank is 0 as claimed.
\end{proof}

\end{document}
